---
title: "class09"
author: "Jason (A15796973)"
format: pdf
---

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

Q2. How many fruity candy types are in the dataset?
```{r}
sum(candy[,2])
```
## Section 2


```{r}
candy["Twix", ]$winpercent
```


Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
## Favorite is Reese's Peanut Butter cup
candy["Reese's Peanut Butter cup", ]$winpercent
```

Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?
```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```


```{r}
library("skimr")
skim(candy)
```


Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The chocolate variable is either 1 or 0, while most rows are a range between 1 and 0.


Q7. What do you think a zero and one represent for the candy$chocolate column?

A 0 means that that the selected candy is not a chocolate type candy, and a 1 menas it is.

Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```


Q9. Is the distribution of winpercent values symmetrical?

```{r}
library(ggplot2)
ggplot(candy, aes(winpercent)) + geom_histogram(binwidth = 10)

## No

## The distribution is not symmetrical
```

Q10. Is the center of the distribution above or below 50%?

Below 50%

Q11. On average is chocolate candy higher or lower ranked than fruit candy?


```{r}
mean( candy[as.logical(candy$chocolate), "winpercent" ] )
mean( candy[as.logical(candy$fruity), "winpercent" ] )

## Chocolate candy is lower ranked than fruit candy
```



Q12. Is this difference statistically significant?
```{r}
t.test(candy[as.logical(candy$chocolate), "winpercent" ], candy[as.logical(candy$fruity), "winpercent" ])
```
It is statistically significant due to the low p-value of 2.871e-08.

## Section 3


Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)
candy %>% arrange(winpercent) %>% head(5)
```


Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>% arrange(winpercent) %>% tail(5)
```

Q15. Make a first barplot of candy ranking based on winpercent values.
```{r}
ggplot(candy, aes(winpercent, rownames(candy)) ) +
  geom_col()
```

Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?
```{r}

ggplot(candy, aes(winpercent, reorder(rownames(candy),winpercent))) +
  geom_col(fill="green")


```



```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```
- Q17. What is the worst ranked chocolate candy?

The worst ranked candy is Nik L Nip.

- Q18. What is the best ranked fruity candy?

The best ranked fruity candy is starburst.


## Section 4

```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 50)
## Need to increase max overlaps to turn it into pdf, normally it is set to 5
```

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )


tail( candy[ord,c(11,12)], n=5 )
```


Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reeses Miniatures

Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, Hershey's Milk Chocolate. The least popular is Nik L Nip.


Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy, aes(winpercent, rownames(candy)) ) +
  geom_col()

ggplot(candy, aes(pricepercent, reorder(rownames(candy),pricepercent))) +
  geom_col()


ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```
## Section 5


```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```


Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity

Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent

## Section 6

```{r}
pca = prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
plot(pca$x[,1:2], col=my_cols, pch=16)
```


```{r}
my_data = cbind(candy, pca$x[,1:3])

p = ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```




```{r}
library(ggrepel)

## Normally 7 max overlaps

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 50)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```



```{r}
library(plotly)
ggplotly(p)
```


```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```


Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

fruity, hard, and pluribus.  This makes sense since fruity candy is often pluribus and there are plenty of hard kinds, while fruity candy is also negatively correlated with chocolate as those two types are almost never found together.
